---
title: "dplyr and tables"
author: "Melinda Higgins"
date: "2024-02-07"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      error = TRUE,
                      warning = FALSE,
                      message = FALSE)

library(dplyr)
```

## Make table of a data.frame

Show a simple data.frame output.

```{r}
mtcars %>%
  select(cyl, mpg, hp) %>%
  arrange(hp) %>%
  filter(hp < 100)
```

Now convert it to a formatted table using `knitr::kable()` function.

```{r}
mtcars %>%
  select(cyl, mpg, hp) %>%
  arrange(hp) %>%
  filter(hp < 100) %>%
  knitr::kable()
```

## Getting summary statistics

```{r}
mtcars %>%
  select(cyl, mpg, hp) %>%
  arrange(hp) %>%
  summarise(meanmpg = mean(mpg),
            meanhp = mean(hp)) %>%
  # use rmarkdown and knitr
  # for nicer rendered table
  knitr::kable(
    col.names = c("MPG","HP"),
    caption = "Means of MPG and HP"
  )
```

## Summary stats with `group_by()`

```{r}
mtcars %>%
  select(cyl, mpg, hp) %>%
  arrange(hp) %>%
  group_by(cyl) %>%
  summarise(meanmpg = mean(mpg),
            meanhp = mean(hp)) %>%
  knitr::kable(
    col.names = c("CYL","MPG","HP"),
    caption = "Means of MPG and HP by CYL")
```

## Use `knitr::kable()` for pretty output table

```{r}
library(palmerpenguins)
penguins %>%
  select(body_mass_g, flipper_length_mm, species) %>%
  group_by(species) %>%
  summarise(
    body_mean = mean(body_mass_g, na.rm=TRUE),
    body_sd = sd(body_mass_g, na.rm=TRUE),
    flip_mean = mean(flipper_length_mm, na.rm=TRUE),
    flip_sd = sd(flipper_length_mm, na.rm=TRUE)
  ) %>%
  knitr::kable()
```

## knitr::kable() - add better column headings

```{r}
penguins %>%
  select(body_mass_g, flipper_length_mm, species) %>%
  group_by(species) %>%
  summarise(
    body_mean = mean(body_mass_g, na.rm=TRUE),
    body_sd = sd(body_mass_g, na.rm=TRUE),
    flip_mean = mean(flipper_length_mm, na.rm=TRUE),
    flip_sd = sd(flipper_length_mm, na.rm=TRUE)
  ) %>%
  knitr::kable(col.names = c("Species",
                             "Mean Body Mass (g)",
                             "SD of Body Mass (g)",
                             "Mean Flipper Length (mm)",
                             "SD of Flipper Length (mm)"))
```

## use `across()` with dplyr `summarise()`

```{r}
penguins %>%
  dplyr::group_by(species) %>%
  dplyr::summarize(across(ends_with("mm"), 
                          ~ mean(.x, na.rm = TRUE))) %>%
  knitr::kable()
```

## Making categories or recoding - dplyr::case_when()

```{r}
# Use case_when() to create
# three size categories
penguins_mod <- penguins %>%
  mutate(size_bin = case_when(
    body_mass_g > 4500 ~ "large",
    body_mass_g > 3000 & 
      body_mass_g <= 4500 ~ "medium",
    body_mass_g <= 3000 ~ "small"
    )
  )
# Create summary table 
# of size categories
penguins_mod %>% 
  pull(size_bin) %>% 
  table(useNA = "ifany") %>%
  knitr::kable(
    col.names = c("Size Category",
                  "Frequency"))
```

## Quick Table Example - Arsenal Package

**NOTE**: When using the `arsenal` package, remember to add `results="asis"` to the R code chunk options. This allows the `arsenal` table to be rendered correctly when you "KNIT" your document.

```{r results = "asis"}
library(arsenal)
tab1 <- tableby(island ~ bill_length_mm + bill_depth_mm, 
                data = penguins)
summary(tab1)
```


## Another quick table with `gtsummary` package

The `gtsummary` package does NOT need `results="asis"` in the code chunk options. And this works with the `dplyr` syntax.

Learn more about `tbl_summary()` and more at the [`gtsummary` package website](https://www.danieldsjoberg.com/gtsummary/articles/tbl_summary.html).

```{r}
library(gtsummary)
penguins %>%
  select(bill_length_mm, bill_depth_mm, island) %>%
  tbl_summary(by = island) %>%
  add_p()
```

